---
title: "EDA and visualisation"
output: html_document
date: "2026-09-21"
---

# EDA

### Missing data choices
The manuscripts with unknown dates or unknown "waves" are removed from the dataset, until further notice.

The data is undirected. Because the directed data assumes that the compiler was only able to choice (agency) a succession of sermons, se1 -> se2 -> se3, but he/she compiles [se1, se2, se3] with knowledge of all three sermons. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load nodes edges, warning=F, echo=F}
root_data_path = "./data/data_deriv/"

se_edge_path = "se_se_undir_080926.xlsx"
se_node_path = "se_nodes_colls_080926.xlsx"

raw_se_data = read_excel("./data/data_raw/sf_mss_okt_25.xlsx") |> 
  drop_na(century)

se_edge <- read_excel(paste0(root_data_path, se_edge_path)) |> 
  drop_na(century)
# 9 manuscripts removed without dating

# some nodes has duplicates and they are collapsed if one has 1 than 1 else 0 to be investigated
se_nodes <- read_excel(paste0(root_data_path, se_node_path)) |> 
  group_by(Id) |> 
  summarise(
    across(everything(), \(x) max(x, na.rm = TRUE)),
    .groups = "drop"
  )

# make raw data century -> int
sf_se_data <- raw_se_data |>
  mutate(century = as.integer(century),
         n_ses = str_count(certain, " ") + 1) |> 
  drop_na()
  
print(head(se_edge, 5))

print(head(se_nodes, 5))
```

## Transpose to igraph

```{r igraph, echo=FALSE}
# edgelist into igraph object
g <- graph_from_data_frame(se_edge, 
                            vertices = se_nodes,
                            directed = F)

# check attibutes
vertex_attr_names(g)
edge_attr_names(g)

```

```{r plot alleluia}
plot_alleluia_century <- function(g_full, cent, delete_non_nodes) {
  # slice graph
  g_i <- subgraph.edges(
    g_full,
    E(g_full)[century == cent],
    delete.vertices = delete_non_nodes
  )
  
  # voeg een weight toe
  E(g_i)$weight = 1
  
  G1 <- simplify(g_i, edge.attr.comb = list(weight = "sum"))
  
  # created weighted edges but normalised between 0 and 1
  E(G1)$weight <- E(G1)$weight / max(E(G1)$weight)
  
  # maak kleuren
  V(G1)$color <- ifelse(
    V(G1)$alleluia_spring == 1,
    "yellow",
    ifelse(V(G1)$alleluia_easter == 1,
           "green",
           ifelse(
              V(G1)$`alleluia_mixed?` == 1,
              "red",
              V(G1)$color <- ifelse(
                V(G1)$alleluia_mixed == 1,
                "orange",
                ifelse(
                  V(G1)$alleluia_after_425 == 1,
                  "blue",
                  "lightgrey"
                )
              ))))
  
  # layout
  layout <- layout_with_mds(G1, dist = NULL, dim = 2)
  # trek uit elkaar
  layout <- norm_coords(layout, xmin = -50, xmax = 50, ymin = -50, ymax = 50)
  # 
  # # node size degree
  V(G1)$size <- degree(G1) / mean_degree(G1) * 300 + 50
  
    # plot century
  p1 <- plot(G1,
    vertex.color = V(G1)$color,
    vertex.label = NA,
    layout=layout,
    rescale=F
  )
  
  return (p1)
}

cents <- seq(8, 13)

par(mfrow = c(2, 3), mar= c(1,1,1,1))

for (c in cents){
  plot_alleluia_century(g_full=g, cent=c, delete_non_nodes=TRUE)
  title(main = paste("Century:", c))
}


```

## Basic
```{r EDA undirected, warning=F}
# get n nodes (removed if degree = 0)
century_count <- function (g_full, cent) {
  # slice graph
  g_i <- subgraph.edges(
    g,
    E(g)[century == cent],
    delete.vertices = T
  )
  
  n_unq_mss = length(unique(E(g_i)$manuscript))
  
  # voeg een weight toe
  E(g_i)$weight = 1
  
  g_i <- simplify(g_i, edge.attr.comb = list(weight = "sum"))
  
  # normalise weight
  E(g_i)$weight <- E(g_i)$weight / max(E(g_i)$weight)
  avg_edges <- if(n_unq_mss > 0) ecount(g_i) / n_unq_mss else 0
  
  return(data.frame(
    Century = cent,
    Nodes = vcount(g_i),
    Manuscripts = n_unq_mss,
    Edges = ecount(g_i),
    Avg_Edge_Per_MS = avg_edges,
    Degree = igraph::mean_degree(g_i),
    Transitivity = igraph::transitivity(g_i, type = "global"),
    Density = igraph::edge_density(g_i),
    stringsAsFactors = FALSE
  ))
}

# Size
# number of nodes
vcount(g)
# number of edges
ecount(g)

centuries <- unique(E(g)$century)

df_results <- do.call(rbind, lapply(centuries, function(c) {
  century_count(g_full = g, cent = c)
}))

knitr::kable(df_results)

sf_se_data |> 
  mutate(century_str = fct_relevel(as.character(century), 
            as.character(seq(4, 16)) 
            )) %>%
  ggplot(aes(x = century, y=n_ses)) +
    geom_smooth() +
    labs(x = "Century", 
         y = "ses/ms (log)", 
         title = "Number of authentic Augustinian sermons per manuscript",
         subtitle = "Centain sermons with an attributed century") +
    scale_y_log10() +
    theme_minimal()

se_edge |> 
  distinct(century, manuscript) |> 
  mutate(century_str = fct_relevel(as.character(century), 
            as.character(seq(4, 16)) 
            )) |> 
  group_by(century_str) |> 
  summarise(mss = n()) |> 
  as.data.frame() |> 
  ggplot(aes(x=century_str, y = mss)) +
  geom_col() +
  theme_minimal() +
  labs(x='Century',
       y = "N manuscripts", 
       title="Number of manuscripts per century")
```

### Type of Collection per century
-> TODO: preek matrix | per century | kleurtjes per subcollectie | Freq Ms 
