---
title: "EDA and visualisation"
output: html_document
date: "2026-09-21"
---

# EDA

### Missing data choices
The manuscripts with unknown dates or unknown "waves" are removed from the dataset, until further notice.

The data is undirected. Because the directed data assumes that the compiler was only able to choice (agency) a succession of sermons, se1 -> se2 -> se3, but he/she compiles [se1, se2, se3] with knowledge of all three sermons. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(readxl)
library(igraph)
```

```{r load nodes edges, warning=F, echo=F}
root_data_path = "./data/data_deriv/"

se_edge_path = "se_se_undir_080926.xlsx"
se_node_path = "se_nodes_colls_080926.xlsx"

raw_se_data = read_excel("./data/data_raw/sf_mss_okt_25.xlsx") |> 
  drop_na(century)

se_edge <- read_excel(paste0(root_data_path, se_edge_path)) |> 
  drop_na(century)
# 9 manuscripts removed without dating

# some nodes has duplicates and they are collapsed if one has 1 than 1 else 0 to be investigated
se_nodes <- read_excel(paste0(root_data_path, se_node_path)) |> 
  group_by(Id) |> 
  summarise(
    across(everything(), \(x) max(x, na.rm = TRUE)),
    .groups = "drop"
  )

# make raw data century -> int
sf_se_data <- raw_se_data |>
  mutate(century = as.integer(century),
         n_ses = str_count(certain, " ") + 1) |> 
  drop_na()
  
print(head(se_edge, 5))

print(head(se_nodes, 5))
```

## Transpose to igraph

```{r igraph, echo=FALSE}
# edgelist into igraph object
g <- graph_from_data_frame(se_edge, 
                            vertices = se_nodes,
                            directed = F)

# check attibutes
vertex_attr_names(g)
edge_attr_names(g)

# prepare legend
incl_colls <- c("Alleluia", "VDA", "Quinquaginta", "Lapsu Mundi", "SCP")
coll_colors <- c("blue", "yellow", "orange", "purple", "darkgreen")

plot(NULL ,xaxt='n',yaxt='n',bty='n',ylab='',xlab='', xlim=0:1, ylim=0:1)
legend("topleft", legend =incl_colls, pch=16, pt.cex=3, cex=1.5, bty='n',
    col = coll_colors)
mtext("Included Collections", at=0.2, cex=2)
```

## Basic
```{r EDA undirected, warning=F}
# get n nodes (removed if degree = 0)
century_count <- function (g_full, cent) {
  # slice graph
  g_i <- subgraph.edges(
    g,
    E(g)[century == cent],
    delete.vertices = T
  )
  
  n_unq_mss = length(unique(E(g_i)$manuscript))
  
  # voeg een weight toe
  E(g_i)$weight = 1
  
  g_i <- simplify(g_i, edge.attr.comb = list(weight = "sum"))
  
  # normalise weight
  E(g_i)$weight <- E(g_i)$weight / max(E(g_i)$weight)
  avg_edges <- if(n_unq_mss > 0) ecount(g_i) / n_unq_mss else 0
  
  return(data.frame(
    Century = cent,
    Nodes = vcount(g_i),
    Manuscripts = n_unq_mss,
    Edges = ecount(g_i),
    Avg_Edge_Per_MS = avg_edges,
    Degree = igraph::mean_degree(g_i),
    Transitivity = igraph::transitivity(g_i, type = "global"),
    Density = igraph::edge_density(g_i),
    stringsAsFactors = FALSE
  ))
}

# Size
# number of nodes
vcount(g)
# number of edges
ecount(g)

centuries <- unique(E(g)$century)

df_results <- do.call(rbind, lapply(centuries, function(c) {
  century_count(g_full = g, cent = c)
}))

knitr::kable(df_results)

sf_se_data |> 
  mutate(century_str = fct_relevel(as.character(century), 
            as.character(seq(4, 16)) 
            )) %>%
  ggplot(aes(x = century, y=n_ses)) +
    geom_smooth() +
    labs(x = "Century", 
         y = "ses/ms (log)", 
         title = "Number of authentic Augustinian sermons per manuscript",
         subtitle = "Centain sermons with an attributed century") +
    scale_y_log10() +
    theme_minimal()

se_edge |> 
  distinct(century, manuscript) |> 
  mutate(century_str = fct_relevel(as.character(century), 
            as.character(seq(4, 16)) 
            )) |> 
  group_by(century_str) |> 
  summarise(mss = n()) |> 
  as.data.frame() |> 
  ggplot(aes(x=century_str, y = mss)) +
  geom_col() +
  theme_minimal() +
  labs(x='Century',
       y = "N manuscripts", 
       title="Number of manuscripts per century")
```

### Conclusion EDA
The number of observations should be more reliabile if it is higher than 50 obervations. In this case, only the 9th to the 15th century can be used for observation and the 5th to the 8th century can be redirected to my qualitative collegues. The 16th century is underrepresented in this dataset, because it only includes manuscripts, which saw a steady decline in production from the 16th century, as well as, discussion about the validity of church fathers within the Reformation debate, resulting in a lower production of augustinian sermons in manuscripts. <b>


```{r plot alleluia}
plot_alleluia_century <- function(g_full, cent, delete_non_nodes) {
  # slice graph
  G1 <- subgraph.edges(
    g_full,
    E(g_full)[century == cent],
    delete.vertices = delete_non_nodes
  )
  
  # maak kleuren
  colour_collection <- ifelse(
    (V(G1)$lapsu == 1) & (cent >= 11), 
            "purple", 
    ifelse(V(G1)$alleluia == 1,
        "blue",
    ifelse(V(G1)$vda == 1, "yellow",
    ifelse(V(G1)$quinquaginta == 1,
           "orange",
    ifelse((V(G1)$scp == 1) & (cent >= 12),
           "darkgreen",
                  "lightgrey"
                )))))
  
  # layout
  layout <- layout_with_mds(G1, dist = NULL, dim = 2)
  # trek uit elkaar
  layout <- norm_coords(layout, xmin = -75, xmax = 75, ymin = -50, ymax = 50)
  # 
  # # node size degree
  V(G1)$size <- betweenness(G1) / max(betweenness(G1)) * 800 + 250
  
  # remove edges but keep layout
  G1_edgeless = delete_edges(G1, edges=E(G1))
  
    # plot century
  p1 <- plot(G1_edgeless,
    vertex.color = adjustcolor(colour_collection, alpha.f = 0.8),
    vertex.label = NA,
    layout=layout,
    rescale=F
  )
  
  return (p1)
}

cents_early <- seq(9, 12)

cens_late <- seq(13, 15)

par(mfrow = c(2, 2), mar= c(1,1,1,1))

for (c in cents_early){
  plot_alleluia_century(g_full=g, cent=c, delete_non_nodes=TRUE)
  title(main = paste("Century:", c))
}

par(mfrow = c(2, 2), mar= c(1,1,1,1))

for (c in cens_late){
  plot_alleluia_century(g_full=g, cent=c, delete_non_nodes=TRUE)
  title(main = paste("Century:", c))
}


# to do:
# misschien geen ties
# alpha = 0.5
# belangrijke collecties toevoegen
# alleluia andere vorm
# plaatje plotten
# betweenness centrality

# standaard -> legend apart, nu niet. 

# global clustering coefficent ~ century

```
### Clustering coefitient
The structure of the networks change over the centuries. In the graph, the centuries after the 12th century display stronger clustering. This can be tested with a global transitivity score, which can be seen in the next plot. 

```{r clustering over centuries}
df_results

```

### Type of Collection per century
-> TODO: preek matrix | per century | kleurtjes per subcollectie | Freq Ms 
